---
title: "Models & Density-independent growth"
author: "Christopher Moore"
date: "Week 02"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: inverse, center, middle
# Introduction to models
---
# Types of models
Form        | Time       | Space              | Predictability 
:----------:|:----------:|:------------------:|:--------------:
Dynamical   | Continuous | Non-spatial        | Deterministic
Verbal      | Discrete   | Spatially-implicit | Stochastic
Graphical   |            | Spatial-explicit   |
Statistical |            |                    |
Etc.        |            |                    |
---
## Continuous and discrete time
### Continous
* Variables in the model change all of the time
* Generations are *overlapping*
* There is no inherent delay between "events"
* No built-in periodicity (e.g., seasonality)

--

### Discrete
* Variables in the model change at predefined, constant increments
* Generations can be *non-overlapping*
* There is a delay between "events"
* Commonly used for periodic (e.g., seasonal) systems
---

## Parts of dynamical models
### State variables: change over time
* Examples in non-ecological models include: concentration, position, temperature, current, allele frequency, GDP
* Examples in ecological models include: abundance, density, position, yield, resource concentration
--

***
### Parameters: constants that generally don't change over time
* Examples in non-ecological models: slection coefficients, distance, viscosity, mutation, decay
* Examples in ecological models: births, deaths, strength of density dependence, sign (e.g., predators always have a - effect on prey)
---

## General anatomy of a dynamical model
### State variable(s)
* For this course, we are focused on abundance or density, commonly abbreviated as $N$
* Time will be abbreviated as $t$

--

### Generally
$$ \frac{\Delta N}{\Delta t} = f(N) $$
--

#### Some conventions
* For discrete-time models we use $N_t$
* For continuous-time models we use $N(t)$. Although for this class, and generally throughout the ecological literature (including in the book), we simply write $N$

---
## Anatomy of a dynamical model: discrete time
* General dyanmic model
$$f(N) = \frac{\Delta N}{\Delta t} = $$
--

* Geometric growth: constant growth, discrete time 
$$N_{t+1} = \lambda N_t$$

--

* Writing a function in `R`
```{r}
geom_growth <- function(Nt, lambda) {
{{Nt1 <- lambda*Nt}}
  return(Nt1)
}
```

--

* An example with $N_t = 1$ and $\lambda = 1.1$:
```{r}
geom_growth(Nt = 1, lambda = 1.1)
```

---
## Anatomy of a dynamical model: continuous time
* General dyanmic model
$$f(N) = \frac{\Delta N}{\Delta t} = $$
--

* Exponential growth: constant growth, continuous time 
$$\frac{\mathrm{d}N}{\mathrm{d}t} = rN$$

--
* Writing a function in `R`
```{r}
exp_growth <- function(Nft, r) {
{{  dNdt <- r*Nft}}
  return(dNdt)
}
```

--
* An example with $N = 1$ and $r = 0.1$:
```{r}
exp_growth(Nft = 1, r = 0.1)
```


---
## Continuous and discrete time (with equations)
### Continous
* Variables in the model change all of the time
* Generations are *overlapping*
* There is no inherent delay between "events"
* No built-in seasonality
$$\frac{\mathrm{d}N}{\mathrm{d}t} = f(N)$$

--

### Discrete
* Variables in the model change at predefined, constant increments
* Generations can be *non-overlapping*
* There is a delay between "events"
* Commonly used for periodic (e.g., seasonal) systems
$$ N_{t+1} = f(N_t) $$
---
class: inverse, center, middle
# Density-indepedent change

---
# The simplest models: constant change
.pull-left[
Geometric growth for $\lambda = 1.1$ and $N_1 = 2$:
$$ \vphantom{\frac{dN}{dt}} \lambda N_t $$
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
nst <- 2
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- nst_vec[i-1] + 0.1*nst_vec[i-1]
}
par(mar = c(4, 4, 0.5, 0.5))
plot(x = steps, y = nst_vec, type = "p", col = "red", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(2, 14))
segments(steps[-length_steps], nst_vec[-length_steps], steps[-1], nst_vec[-length_steps], co = "red", lwd = 1.5)
```
]

--

.pull-right[
Exponential growth for $r = 0.1$ and $N(1) = 2$:
$$ \frac{\mathrm{d}N}{\mathrm{d}t} = rN $$

```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
n0 <- 2
r <- 0.1
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
plot(x = (times++1), y = nt, type = "l", col = "blue", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(2, 14), lwd = 1.5)
```
]

---
# Note that continous time is always greater than discrete time
```{r, echo = F, out.width = 400, out.height = 400, dpi = 400, fig.width = 4, fig.height = 4, fig.align="center"}
nst <- 2
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- nst_vec[i-1] + 0.1*nst_vec[i-1]
}
n0 <- 2
r <- 0.1
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
plot(x = (times+1), y = nt, type = "l", col = "blue", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(2, 14), lwd = 1.5)
segments(steps[-length_steps], nst_vec[-length_steps], steps[-1], nst_vec[-length_steps], col = "red", lwd = 1.5)
points(x = steps, y = nst_vec, pch = 16, cex = 0.75, col = "red")
```

---
# The simplest models: constant change
.pull-left[
Geometric growth for $\lambda = 0.75$ and $N_1 = 10$:
$$ \vphantom{\frac{dN}{dt}} N_{t+1} = \lambda N_t $$
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
nst <- 10
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- 0.75*nst_vec[i-1]
}
par(mar = c(4, 4, 0.5, 0.5))
plot(x = steps, y = nst_vec, type = "p", col = "red", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(0, 10))
segments(steps[-length_steps], nst_vec[-length_steps], steps[-1], nst_vec[-length_steps], co = "red", lwd = 1.5)
segments(-5, 0, 25, 0)
```
]

--

.pull-right[
Exponential growth for $r = -0.25$ and $N(1) = 10$:
$$ \frac{\mathrm{d}N}{\mathrm{d}t} = rN $$

```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
n0 <- 10
r <- -0.25
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
plot(x = (times++1), y = nt, type = "l", col = "blue", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(0, 10), lwd = 1.5)
segments(-5, 0, 25, 0)
```
]
---
# Note that continous time is always greater than discrete time
```{r, echo = F, out.width = 400, out.height = 400, dpi = 400, fig.width = 4, fig.height = 4, fig.align="center"}
nst <- 10
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- nst_vec[i-1] - 0.25*nst_vec[i-1]
}
n0 <- 10
r <- -0.25
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
plot(x = (times+1), y = nt, type = "l", col = "blue", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(0, 10), lwd = 1.5)
segments(steps[-length_steps], nst_vec[-length_steps], steps[-1], nst_vec[-length_steps], col = "red", lwd = 1.5)
points(x = steps, y = nst_vec, pch = 16, cex = 0.75, col = "red")
segments(-5, 0, 25, 0)
```
---
class: inverse, center, middle
# So I have a model, what do I do with it?

---
# What do to with models <small>(including density-independent growth and everything we wil cover for this course)</small>

* Analyze it!
    * Simulate it
    * Solve it
    * Find long-term behavior
    * Determine behavior if perturbed
    * Make predictions
  
--

* Vizualize it!
    * Plot a time series of $t$ and population size ( $N$ )
    * Plot population size ( $N$ ) and population growth $\frac{\Delta N}{\Delta t}$
    * Plot population size ( $N$ ) and growth per individual (per-capita) $\frac{1}{N}\frac{\Delta N}{\Delta t}$
    
---
# Analyze it!: simulation
.pull-left[
**Code and output** for discrete-time growth <small>(simulating it)</small>:
```{r, collapse = T}
max_time <- 20
timesteps <- 1:100
N <- vector(mode = "numeric", length = max_time)
N[1] <- 2
lambda <- 1.1
for (i in 2:max_time) {
  N[i] <- lambda*N[i-1]
}
round(N[1:3], 2)
round(N[(max_time - 2):max_time], 2)
```
]

.pull-right[
**Code and output** for continuous-time growth <small>(not simulating it, but using the solution [simulating differential equations is a bit more involved])</small>:
```{r, collapse = T}
N0 <- 2
r <- 0.1
timesteps <- seq(0, 20, by = 1)
len_time <- length(timesteps)
N <- N0*exp(r*timesteps)
round(N[1:3], 2)
round(N[(len_time - 2):len_time], 2)
```
]

---
# Analyze it!: solve it

.pull-left[
**Discrete time**:
$$\begin{align}
N_{t+1} &= \lambda N_t \\
N_{t+2} &= \lambda N_{t+1} \\
N_{t+2} &= \lambda (\lambda N_t) \\
N_{t+2} &= \lambda ^2 N_t \\
N_{t+3} &= \lambda ^3 N_t \\
... \\
N_{t} &= \lambda ^t N_0
\end{align}$$
]
--
.pull-right[
**Continuous time**:
$$\begin{align}
\frac{\mathrm{d}N}{\mathrm{d}t} &= rN \\
\frac{1}{N}\mathrm{d}N &= r \mathrm{d}t \\
\int_{N(0)}^{N(T)} \frac{1}{N}\mathrm{d}N &= \int_{0}^{T} r \mathrm{d}t \\
\log _e(N)\Big|_{N(0)}^{N(T)} &= rt\Big|_0^T \\
\log _e\left(\frac{N(T)}{N(0)}\right) &= rT \\
N(T) &= N(0)e^{rT}
\end{align}$$
]

---
# Analyze it!: Find long-term behavior, determine behavior if perturbed, make predictions
Not too much happening with this simple, linear model:
* Long-term behavior: continues to grow or shrink at a constant rate
* Determinted behavior if perturbed: continues to grow or shrink at a constant rate
* Make predictions: continues to grow or shrink at a constant rate
---
# Vizualize it!: Plot a time series <small>(already did it)</small>
.pull-left[
Geometric growth for $\lambda = 1.1$ and $N_1 = 2$:
$$ \vphantom{\frac{\mathrm{d}N}{\mathrm{d}t}} N_{t+1} = \lambda N_t $$
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
nst <- 2
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- 1.1*nst_vec[i-1]
}
par(mar = c(4, 4, 0.5, 0.5))
plot(x = steps, y = nst_vec, type = "p", col = "red", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(0, 14))
segments(steps[-length_steps], nst_vec[-length_steps], steps[-1], nst_vec[-length_steps], col = "red", lwd = 1.5)
```
]

--

.pull-right[
Exponential growth for $r = 0.1$ and $N(1) = 2$:
$$ \frac{\mathrm{d}N}{\mathrm{d}t} = rN $$

```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
n0 <- 2
r <- 0.1
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
plot(x = (times+1), y = nt, type = "l", col = "blue", las = 1, xlab = bquote(italic(t)), ylab = bquote("Density"~(italic(N))), pch = 16, cex = 0.75, xlim = c(0, 21), ylim = c(0, 14), lwd = 1.5)
```
]
---
# Vizualize it!: Plot population growth against density
.pull-left[
Geometric growth for $\lambda = 1.1$ and $N_1 = 2$:
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
nst <- 2
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- nst_vec[i-1] + 0.1*nst_vec[i-1]
}
par(mar = c(4, 4, 0.5, 0.5))
nt <- nst_vec[1:(length(nst_vec)-1)]
nt1 <- nst_vec[2:length(nst_vec)]
diffnt <- nt1-nt
plot(x = nt, y = diffnt, type = "p", col = "red", las = 1, xlab = bquote("Density at time t"~(italic(N[t]))), ylab = "Difference of density at N_{t+1} and N_{t}", pch = 16, cex = 0.75, xlim = c(0, max(nt)), ylim = c(0, max(diffnt)))
```
]
--
.pull-right[
Exponential growth for $r = 0.1$ and $N(1) = 2$:
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
n0 <- 2
r <- 0.1
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
nT <- nt[1:(length(nt)-1)]
nT1 <- nt[2:length(nt)]
diffnT <- nT1 - nT
plot(x = nT, y = diffnT, type = "l", col = "blue", las = 1, xlab = bquote("Density at time t"~(italic(N(0)))), ylab = "Approaching the derivative", pch = 16, cex = 0.75, xlim = c(0, max(nT)), ylim =c(0, max(diffnT)), lwd = 1.5)
```
]
---
# Vizualize it!: Plot per-capita growth against density
.pull-left[
Geometric growth for $\lambda = 1.1$ and $N_1 = 2$:
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
nst <- 2
length_steps <- 20
steps <- 1:length_steps
nst_vec <- vector(mode = "numeric", length = length_steps)
nst_vec[1] <- nst
for (i in 2:(length(nst_vec))) {
  nst_vec[i] <- 1.1*nst_vec[i-1]
}
par(mar = c(4, 4, 0.5, 0.5))
nt <- nst_vec[1:(length(nst_vec)-1)]
nt1 <- nst_vec[2:length(nst_vec)]
diffnt <- nt1-nt
pcg <- diffnt/nt
plot(x = nt, y = pcg, type = "p", col = "red", las = 1, xlab = bquote("Density at time t"~(italic(N[t]))), ylab = "Per-capita growth", pch = 16, cex = 0.75, xlim = c(0, max(nt)), ylim = c(0, max(pcg)))
```
]
--
.pull-right[
Exponential growth for $r = 0.1$ and $N(1) = 2$:
```{r, echo = F, out.width = 300, out.height = 300, dpi = 400, fig.width = 4, fig.height = 4}
n0 <- 2
r <- 0.1
times <- seq(0, (length_steps-1), by = 0.001)
nt <- n0*exp(r*times)
par(mar = c(4, 4, 0.5, 0.5))
nT <- nt[1:(length(nt)-1)]
nT1 <- nt[2:length(nt)]
diffnT <- nT1 - nT
pcG <- diffnT/nT
plot(x = nT, y = pcG, type = "l", col = "blue", las = 1, xlab = bquote("Density at time t"~(italic(N(0)))), ylab = "Per-capita growth rate", pch = 16, cex = 0.75, xlim = c(0, max(nT)), ylim =c(0, max(pcG)), lwd = 1.5)
```
]


---
# A note about an assumption of *r*
### $\lambda$ and $r$ are the difference between births and deaths

--

Specific to our discrete-time density-independent models, this means:
$$N_{t+1} = \lambda N_t = \left(B - D\right)N_t = BN_t - DN_t$$

--

And the solution would be:
$$N_t = N_0(B - D)^t$$

--

Specific to our continous-time density-independent models, this means:
$$\frac{\mathrm{d}N}{\mathrm{d}t} = rN = (b-d)N$$

--

And the solution would be:
$$N(t) = N(0)e^{t(b - d)}$$